---
title: "SolNRCH_foldome_v1"
author: "AmirAli Toghani"
date: "2025-12-11"
output: html_document
editor_options:
  chunk_output_type: console
---

<!--
File-path conventions used in this document:
  ../supplementary/       paper supplementary data (Data_S1.xlsx etc.)
  ../phylo/               phylogenetic analysis inputs/outputs
  ../structural_analysis/ motif coordinate tables consumed by resistosome_pipeline
  /path/to/...            external inputs/outputs that live outside this repository
                          (NLRtracker output, NLRexpress output, CD-HIT output,
                          reference FASTA files, AlphaFold 3 input/output). Replace
                          these placeholders with local paths before running.
-->

```{r}
library(tidyverse)
library(readxl)
library(Biostrings)
library(ggtree)
library(pheatmap)
library(reshape2)
library(svglite)
```

## 1. import the species/genome metadata and reference sequences from Solanaceae genomes:
```{r}
metadata <- read_xlsx("../supplementary/Data_extS1.xlsx")

species_meta <- metadata[,c(2,3,4,5,6,14,20)]

NRC_ref <- readAAStringSet("../supplementary/NRC_ref_clu.fasta")
NRC_ref_NBARC <- readAAStringSet("../supplementary/NRC_ref_clu_NBARC.fasta")

MHD <- NLRtracker[NLRtracker$description == "MHD",]
Ploop <- NLRtracker[NLRtracker$description == "P-loop",]

species_freq <- table(species_meta$Species) %>% as.data.frame()
```


## 2. Import sequences and metadata:
```{r}
setwd("/path/to/nlrtracker")

nlrtracker_files <- list.files(pattern = "_Domains.tsv", recursive = T)

nlrtracker_list <- list()

# Loop through each file
for (i in seq_along(nlrtracker_files)) {
  # Read the Excel file
  nlrtracker_list[[i]] <- read_delim(nlrtracker_files[i])

  # Extract the specific part of the file name and add it as a new column
  nlrtracker_list[[i]] <- nlrtracker_list[[i]] %>%
    mutate('Proteome File Name' = sub("_Domains\\.tsv$", ".gz", sub("^nlrtracker_", "", basename(nlrtracker_files[i]))))

  # Print a message for each file (optional)
  cat("Read file:", nlrtracker_files[i], "and added formatted filename column\n")
}


# merge all and add the metadata
NLRtracker <- do.call(rbind,nlrtracker_list) %>% inner_join(species_meta, by = "Proteome File Name")


NLRtracker <- NLRtracker %>%
  mutate(ID = paste(`Assembly Accession`, Species, seqname, Simple, sep = "_")) %>%
  # Replace spaces in the "Species" column with underscores for the "ID" column
  mutate(ID = gsub(" ", "_", ID))

```

## 3. Prepare the domain prerequisites:
```{r}

domains <- c("CNL","CNLO","CN","OCNL","CONL",
             "NL","NLO","ONL",
             "BCNL","BCNLO","BNL","BBNL","BCN","BCCNL","BBCNL","BNLO","BOCNL","BBCNLO",
             "RNL",
             "TN","TNL","TNLO","TNLJ")

```


## 4. Extract the NLRs and NBARCs:
```{r}
NLR <- filter(NLRtracker, NLRtracker$type == "CHAIN" & NLRtracker$Status == "NLR")

# only keep the ones with desired domain architechture
NLR_filtered <- NLR[NLR$Simple %in% domains,]

NLR_filtered_seq <- AAStringSet(NLR_filtered$sequence)
NLR_filtered_seq@ranges@NAMES <- NLR_filtered$ID

# remove duplicate sequences
NLR_filtered_deduplicated_seq <- readAAStringSet("/path/to/NLR_filtered_seq_clu.fasta")

NLR_filtered_deduplicated <- NLR_filtered[NLR_filtered$ID %in% NLR_filtered_deduplicated_seq@ranges@NAMES,]

# extract the NBARC domains
NBARC <- NLRtracker[NLRtracker$description == "NBARC",]
write_csv(NBARC, "/path/to/NBARC.csv")

NBARC_filtered <- NBARC[NBARC$ID %in% NLR_filtered$ID,]
NBARC_filtered_seq <- AAStringSet(NBARC_filtered$sequence)
NBARC_filtered_seq@ranges@NAMES <- NBARC_filtered$ID

NBARC_filtered_deduplicated <- NBARC[NBARC$ID %in% NLR_filtered_deduplicated$ID,]

# filter out NLRs with truncated NBARC domain
NBARC_filtered_deduplicated_len <- filter(NBARC_filtered_deduplicated,
                                                        NBARC_filtered_deduplicated$end -
                                                        NBARC_filtered_deduplicated$start > 250 &
                                                          NBARC_filtered_deduplicated$end -
                                                          NBARC_filtered_deduplicated$start < 400) # NBARCs



NLR_filtered_deduplicated_len <- NLR_filtered_deduplicated[NLR_filtered_deduplicated$ID %in% NBARC_filtered_deduplicated_len$ID,] # NLRs


# convert the final data to biostring objects
NLR_filtered_deduplicated_len_seq <- AAStringSet(NLR_filtered_deduplicated_len$sequence)
NLR_filtered_deduplicated_len_seq@ranges@NAMES <- NLR_filtered_deduplicated_len$ID

NBARC_filtered_deduplicated_len_seq <- AAStringSet(NBARC_filtered_deduplicated_len$sequence)
NBARC_filtered_deduplicated_len_seq@ranges@NAMES <- NBARC_filtered_deduplicated_len$ID

```

### 5. Import the NRC-H clade and extract the sequences for phylogenetic analysis:
```{r}
NRCH_clade_tree <- read.tree("../phylo/NRCH_clade.tree")

NRCH_clade <- NLR_filtered_deduplicated_len[NLR_filtered_deduplicated_len$ID %in% NRCH_clade_tree$tip.label,]
NRCH_clade_NBARC <- NBARC_filtered_deduplicated_len[NBARC_filtered_deduplicated_len$ID %in% NRCH_clade_tree$tip.label,]
NRCH_clade_NBARC_withCC <- NRCH_clade_NBARC[NRCH_clade_NBARC$start > 100,] # make sure there's a CC domain


NRCH_clade_filtered <- NRCH_clade[NRCH_clade$Simple %in% c("CNL","CONL","NL","NLO"),] # get rid of IDs and potential misanotations
NRCH_clade_filtered <- NRCH_clade_filtered[NRCH_clade_filtered$ID %in% NRCH_clade_NBARC_withCC$ID,]



NRCH_clade_filtered_len <- NRCH_clade_filtered[NRCH_clade_filtered$end > 800,] # length filter
NRCH_clade_filtered_len_NBARC <- NRCH_clade_NBARC_withCC[match(NRCH_clade_filtered_len$ID, NRCH_clade_NBARC_withCC$ID),]



# convert into biostring format
NRCH_clade_filtered_len_seq <- AAStringSet(NRCH_clade_filtered_len$sequence)
NRCH_clade_filtered_len_seq@ranges@NAMES <- NRCH_clade_filtered_len$ID
writeXStringSet(NRCH_clade_filtered_len_seq, "../phylo/NRCH_clade_filtered_len_seq.fasta")


NRCH_clade_filtered_len_NBARC_seq <- AAStringSet(NRCH_clade_filtered_len_NBARC$sequence)
NRCH_clade_filtered_len_NBARC_seq@ranges@NAMES <- NRCH_clade_filtered_len_NBARC$ID


```

### 6. Extract the CC-NBARC portion and prepare for AF3:
```{r}

NRCH_clade_filtered_len_seq_CCNBARC <- subseq(NRCH_clade_filtered_len_seq, start = 1, end = NRCH_clade_filtered_len_NBARC$end)

NRCH_clade_filtered_len_seq_CCNBARC_filtered <- NRCH_clade_filtered_len_seq_CCNBARC[NRCH_clade_filtered_len_seq_CCNBARC@ranges@width < 600]

writeXStringSet(NRCH_clade_filtered_len_seq_CCNBARC_filtered, "/path/to/AF3/NRCH_clade_filtered_len_seq_CCNBARC_filtered.fasta")


NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC <- NRCH_clade_filtered_len_NBARC_seq[NRCH_clade_filtered_len_seq_CCNBARC_filtered@ranges@NAMES]
NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC_ref <- c(NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC, NRC_ref_NBARC)
writeXStringSet(NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC_ref, "../phylo/NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC.fasta")

```


### 7. Additional filtering step to keep sequences with 95% similarity threshold:
```{r}

NRCH_clade_filtered_len_seq_filtered_95 <- readAAStringSet("../phylo/NRCH_clade_filtered_len_seq_95.fasta")

NRCH_clade_filtered_len_filtered_95 <- NRCH_clade_filtered_len[NRCH_clade_filtered_len$ID %in% NRCH_clade_filtered_len_seq_filtered_95@ranges@NAMES,]

write(NRCH_clade_filtered_len_filtered_95$ID, "../phylo/NRCH_clade_filtered_len_seq_95_ids.txt")


NRCH_clade_filtered_len_seq_CCNBARC_filtered_95 <- NRCH_clade_filtered_len_seq_CCNBARC_filtered[NRCH_clade_filtered_len_seq_CCNBARC_filtered@ranges@NAMES %in% NRCH_clade_filtered_len_seq_filtered_95@ranges@NAMES]


NRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC <- NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC[NRCH_clade_filtered_len_seq_CCNBARC_filtered_NBARC@ranges@NAMES %in% NRCH_clade_filtered_len_seq_filtered_95@ranges@NAMES]
writeXStringSet(NRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC, "../phylo/NRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC.fasta")

NRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC_ref <- c(NRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC, NRC_ref_NBARC)
writeXStringSet(NRNRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC_ref, "../phylo/NRCH_clade_filtered_len_seq_CCNBARC_filtered_95_NBARC_ref.fasta")

# full length NLRs from the final 637 set
NRCH_clade_filtered_len_seq_filtered_95_full <- NRCH_clade_filtered_len_seq[NRCH_clade_filtered_len_seq_CCNBARC_filtered_95@ranges@NAMES]
writeXStringSet(NRCH_clade_filtered_len_seq_filtered_95_full, "../phylo/NRCH_clade_filtered_len_seq_filtered_95_full.fasta")

NRCH_clade_filtered_len_filtered_95_final <- NRCH_clade_filtered_len_filtered_95[NRCH_clade_filtered_len_filtered_95$ID %in% NRCH_clade_filtered_len_seq_filtered_95_full@ranges@NAMES,]
write_csv(NRCH_clade_filtered_len_filtered_95_final, "../supplementary/Data_S3.csv")



# extract the NB-ARC domain metadata for the final set
NRCH_clade_filtered_len_NBARC_filtered_95 <- NRCH_clade_filtered_len_NBARC[NRCH_clade_filtered_len_NBARC$ID %in% NRCH_clade_filtered_len_seq_CCNBARC_filtered_95@ranges@NAMES,]
write_csv(NRCH_clade_filtered_len_NBARC_filtered_95, "../supplementary/Data_S3.csv")

NRCH_clade_filtered_len_NBARC_filtered_95_coords <- NRCH_clade_filtered_len_NBARC_filtered_95[,c(18,4,5)] %>% left_join(NRCH_ids_filtered, by = "ID")
write_csv(NRCH_clade_filtered_len_NBARC_filtered_95_coords, "../structural_analysis/main/NRCH_clade_filtered_len_NBARC_filtered_95_coords.csv")


```


## Extract the MHD and P-loop information:
```{r}
MHD_filtered <- MHD[MHD$ID %in% NRCH_ids_filtered$ID,]

Ploop_filtered <- Ploop[Ploop$ID %in% NRCH_ids_filtered$ID,]

```

## Since many MHD and P-loop annotations was missing, import NLRexpress output:
```{r}
# For reference sequences
reference_ids <- tibble(ID = c("M","NbNRC2a","NbNRC4c","NRC3","RPP1-WsA","Sr35","Wai3","ZAR1"),
                        ID_normalized = c("m","nbnrc2a","nbnrc4c","nrc3","rpp1_wsa","sr35","wai3","zar1"))


references_nlrexp <- read_tsv("/path/to/NLRexpress/references/compiled_motifs_top_per_motif.tsv") %>% left_join(reference_ids, join_by("ProtName" == "ID"))

references_MHD_nlrexp <- references_nlrexp[references_nlrexp$Motif == "MHD",]
references_MHD_nlrexp_cooreds <- references_MHD_nlrexp[,c(11,8,9)]
write_csv(references_MHD_nlrexp_cooreds, "/path/to/references_MHD_nlrexp_cooreds.csv")

references_ploop_nlrexp <- references_nlrexp[references_nlrexp$Motif == "P-loop",]
references_ploop_nlrexp_cooreds <- references_ploop_nlrexp[,c(11,8,9)]
write_csv(references_ploop_nlrexp_cooreds, "/path/to/references_ploop_nlrexp_cooreds.csv")


# For NRC helpers
NRCH_nlrexp <- read_tsv("/path/to/NLRexpress/NRCH_clade_filtered_len_seq_filtered_95_full/compiled_motifs_top_per_motif.tsv")

NRCH_MHD_nlrexp <- NRCH_nlrexp[NRCH_nlrexp$Motif == "MHD",] %>% left_join(NRCH_ids_filtered, join_by("ProtName" == "ID"))

NRCH_MHD_nlrexp_coords <- NRCH_MHD_nlrexp[,c(11,8,9)]
write_csv(NRCH_MHD_nlrexp_coords, "../structural_analysis/main/NRCH_MHD_nlrexp_coords.csv")

NRCH_ploop_nlrexp <- NRCH_nlrexp[NRCH_nlrexp$Motif == "P-loop",] %>% left_join(NRCH_ids_filtered, join_by("ProtName" == "ID"))

NRCH_ploop_nlrexp_coords <- NRCH_ploop_nlrexp[,c(11,8,9)]
write_csv(NRCH_ploop_nlrexp_coords, "../structural_analysis/main/NRCH_ploop_nlrexp_coords.csv")



# check how many are missing and if they can be recovered fro NLRtracker output
NRCH_ploop_nlrexp_missing <- NRCH_nlrexp[!NRCH_nlrexp$ProtName %in% NRCH_ploop_nlrexp$ProtName,]

NRCH_ploop_complement <- Ploop_filtered[Ploop_filtered$ID %in% NRCH_ploop_nlrexp_missing$ProtName,]
```


